---
title: "Reproducible Research: Peer Assessment 2"
output: 
  html_document:
    keep_md: true
---


# **Title**

## Types of weather events that are most harmful with respect to population health and with respect to economic consequences

# **Synopsis**


# **Data Processing**


## Loading and preprocessing the data
  
### 1. Downloading/loading "Storm Data" file and reading the CSV file

```{r  echo=TRUE, cache=TRUE}
# suppressPackageStartupMessages(library(data.table))

## I need to change my LOCALE (spanish) to English...
Sys.setlocale('LC_TIME', 'en_US.UTF-8')

compressedFileName <- "repdata%2Fdata_StormData.csv.bz2"

URL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"


# If file doesn't exist in current working directory, I try to download from website
if(!file.exists(compressedFileName))
    download.file(URL, compressedFileName)

# If the file doesn't exist in current working directory after downloading, there are any
# problem downloading this file, or there are any other problem...aborting
if(!file.exists(compressedFileName))    
    stop("Required file doesn't exist in the current working directory! Aborting...")
    
# loading data
data <- read.csv(compressedFileName)

# Get the column names and check if they are properly formatted
colnames(data) <- make.names(colnames(data))

```

```{r echo=TRUE}

head(data)
```

### 2. Storm Data - Exploratory analysis 

There is some documentation of this dataset available [here](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf). 
More information can be obtained from the [NOAA Storm Events Database](http://www.ncdc.noaa.gov/stormevents/ftp.jsp). 

There are **`r nrow(data)` rows** or observations and **`r ncol(data)` columns** or variables in this dataset.

Reading these two websites above and, for the topic of this paper, we can subset the *data* CSV file extracting only the following columns or variables:   

* BGN_DATE (event start date)
* STATE (the USA state where event occurs)
* EVTYPE (climatic event)
* FATALITIES (numerical value for an occurrence of death)
* INJURIES (numerical value for non fatal damage, hurt, harm, ...)
* PROPDMG (numerical value for property damage)
* PROPDMGEXP (value for adjusting PROPDMG value)
* CROPDMG (numerical value for crop damage)
* CROPDMGEXP (value for adjusting CROPDMG value)

The others columns or variables aren't important for the analysis in this article.

Now, we are going to subset the dataset.

```{r echo=TRUE, cache=TRUE, message=FALSE, warning=FALSE}
suppressPackageStartupMessages(library(dplyr))

# select only the variables we are interested in
sub_data <- data %>% select(BGN_DATE, STATE, EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP,CROPDMG, CROPDMGEXP)

# For saving computer resources, we can remove the initial dataset
rm(data)

# now, we have a smaller dataset
head(sub_data)
```

For this analysis we want to study types of weather events that are most harmful with respect to population health and with respect to economic consequences, so we want dataset observations with values over "0" (zero) in the following variables: FATALITIES, INJURIES, PROPDMG and CROPDMG.

```{r echo=TRUE, cache=TRUE}

sub_data <- sub_data %>% filter(FATALITIES > 0 | INJURIES > 0 | PROPDMG > 0 | CROPDMG > 0)

```

Now, there are **`r nrow(sub_data)` rows** or observations and **`r ncol(sub_data)` columns** or variables in *sub_data* dataset. We have reduced the data size significantly.

```{r echo=TRUE, cache=TRUE}
#summary of our dataset
summary(sub_data)

```

### 3. Storm Data - Cleaning the Data

In the official documentation, [here](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf), we can read that there are 48 event types (chapter 7, points from 7.1 to 7.48), and we have `r length(levels(sub_data$EVTYPE))` event types in our dataset. We'll need to transform and reduce them.

In first place, we are going to know how many ocurrences of each event type there are in our *sub_data* dataset.

```{r echo=TRUE, cache=TRUE}
summary(sub_data$EVTYPE)
```

In this summary, we can see that THUNDERSTORM WIND and other variations of the same climatic event are the more common (TSTM WIND, THUNDERSTORM WINDS, THUNDERSTORM WINDSS, THUNDERSTORM, TSTM WIND (G45), THUNDERSTORM WINDS/HAIL, WINDS, GUSTY WINDS). The idea is to group these events in one big group with name **THUNDERSTORM WIND**, like this event is named in the official documentation. We'll do the same process for others climatic events with similar variable name in our dataset. This way we reduce the dataset dimensions.

Also, there are duplicate event names, and the same event name in lowercase and in uppercase, so we'll transform all event names to uppercase, for grouping purpouses.

```{r echo=TRUE, cache=TRUE}

# Transform all event names to UPPERCASE. This transform the column "as.character".
sub_data <- mutate(sub_data, EVTYPE = as.factor(toupper(EVTYPE)))

```


We are going to group similar climatic events that can be repetitive or similar events, like THUNDERSTORM WIND or TSTM WIND, for reduce this dataset. We start looking for and replacing the more harmful climatic events like TORNADO, HURRICANES, ...

```{r echo=TRUE, cache=TRUE}
# searching by TORNADO
toMatch <- c("TORNADO")

# get the indices where there are matches
indices <- grep(paste(toMatch, collapse="|"), sub_data_health$EVTYPE)

# replacement matches with TORNADO
sub_data_health$EVTYPE[indices] = "TORNADO"

```

We are going to do the above same process with others event types. For this purpouse, we are going to create a function to do this.

```{r echo=TRUE, cache=TRUE, message=FALSE, warning=FALSE}


toMatch <-c("TORNADO", "TORNADO")
toMatch <-c("THUNDERST", "WIND", "TSTM", "THUNDERSTORM WIND")
toMatch <-c("HURRICANE", "TYPHOON", "TROPICAL", "HURRICANE/TYPHOON")
toMatch <-c("FLOOD", "FLOOD")
toMatch <-c("SNOW", "SNOW")
toMatch <-c("RAIN", "RAIN")
toMatch <-c("COLD","HYPOTHERMIA/EXPOSURE", "LOW TEMPERATURE", "FREEZE", 
            "FREEZING SPRAY", "COLD")
toMatch <-c("LIGHTNING", "LIGHTNING")
toMatch <-c("FOG", "FOG")
toMatch <-c("WATER", "WATERSPOUT")
toMatch <-c("COASTAL", "COASTAL FLOOD")
toMatch <-c("ICE", "SMALL HAIL", "ICE")
toMatch <-c("HOT", "WARM", "HEAT", "HEAT")
toMatch <-c("DROUGHT", "DROWNING", "DROUGHT")

```




Now, we have two different questions to resolve:

> 1. Weather events that are most harmful with respect to population health
> 2. Weather events that are most harmful with respect to economic consequences

The climatic events that affect the first question may not be the same that affect the second question and vice versa.

For this purpouse, we need to extract these differents observations from the *sub_data* dataset. One dataset for population health, and another dataset for economic consequences.

```{r echo=TRUE, cache=TRUE}
sub_data_health <- sub_data %>% filter(FATALITIES > 0 | INJURIES > 0)
sub_data_economic <- sub_data %>% filter(PROPDMG > 0 | CROPDMG > 0)

# For saving computer resources, we can remove the initial dataset
rm(sub_data)
```

Now, there are **`r nrow(sub_data_health)` rows** or observations and **`r ncol(sub_data_health)` columns** or variables in *sub_data_health* dataset, and  **`r nrow(sub_data_economic)` rows** or observations and **`r ncol(sub_data_economic)` columns** or variables in *sub_data_economic* dataset.

We are going to group similar climatic events that can be repetitive or similar events, like THUNDERSTORM WIND or TSTM WIND, for reduce this dataset. We start looking for and replacing the more harmful climatic events like TORNADO, HURRICANES, ...

```{r echo=TRUE, cache=TRUE}
# searching by TORNADO
toMatch <- c("TORNADO")

# get the indices where there are matches
indices <- grep(paste(toMatch, collapse="|"), sub_data_health$EVTYPE)

# replacement matches with TORNADO
sub_data_health$EVTYPE[indices] = "TORNADO"

```

We are going to do the above same process with others event types. For this purpouse, we are going to create a function to do this.

```{r echo=TRUE, cache=TRUE, message=FALSE, warning=FALSE}


toMatch <-c("TORNADO", "TORNADO")
toMatch <-c("THUNDERST", "WIND", "TSTM", "THUNDERSTORM WIND")
toMatch <-c("HURRICANE", "TYPHOON", "TROPICAL", "HURRICANE/TYPHOON")
toMatch <-c("FLOOD", "FLOOD")
toMatch <-c("SNOW", "SNOW")
toMatch <-c("RAIN", "RAIN")
toMatch <-c("COLD","HYPOTHERMIA/EXPOSURE", "LOW TEMPERATURE", "FREEZE", 
            "FREEZING SPRAY", "COLD")
toMatch <-c("LIGHTNING", "LIGHTNING")
toMatch <-c("FOG", "FOG")
toMatch <-c("WATER", "WATERSPOUT")
toMatch <-c("COASTAL", "COASTAL FLOOD")
toMatch <-c("ICE", "SMALL HAIL", "ICE")
toMatch <-c("HOT", "WARM", "HEAT", "HEAT")
toMatch <-c("DROUGHT", "DROWNING", "DROUGHT")

```


At this point, we are going to check the more harmful event types in our datasets.

```{r echo=TRUE, cache=TRUE}

# Group by event type
my_health_group <- group_by(sub_data_health, EVTYPE)

# Get the sum of injuries and fatalities by event type
my_health_summary <- summarise(my_health_group, sum(FATALITIES), sum(INJURIES))

# change tne column names
colnames(my_health_summary) <- c("EVTYPE", "SUM_FATALITIES", "SUM_INJURIES")

# Order the dataset by number of FATALITIES & INJURIES
my_health_summary <- arrange(my_health_summary, desc(SUM_FATALITIES, SUM_INJURIES))

# Preview
head(my_health_summary)

```

Now we have `r length(levels(my_health_summary$EVTYPE))` different event types in our *my_health_summary* dataset. 


levels(as.factor(grep("*Thunderst*", sub_data$EVTYPE, ignore.case = T, value = T)))
my_group <- group_by(sub_data,EVTYPE, INJURIES, FATALITIES, PROPDMG, CROPDMG)
summarise_each(my_group, funs(sum))