---
title: "Reproducible Research: Peer Assessment 2"
output: 
  html_document:
    keep_md: true
    geometry: margin=1in
---


# **Title**

## Types of weather events that are most harmful with respect to population health and with respect to economic consequences

# **Synopsis**

In this article we want to study what types of weather events that are most harmful with respect to population health and with respect to economic consequences. First, we'll download the NOAA Storm Data CSV file from his website for processing. This processing involves an initial exploratory analysis that checks what variables are more relevant for our study, cleaning the data and grouping the similar events types to get a more uniform dataset. Then, we'll do a summary of our health and economic datasets and present the results showing the datasets and plotting them. Finally, we'll present a final conclusion for our health and economic results.

# **Data Processing**


## Loading and preprocessing the data
  
### 1. Downloading/loading "Storm Data" file and reading the CSV file

```{r  echo=TRUE, cache=TRUE}
# suppressPackageStartupMessages(library(data.table))

## I need to change my LOCALE (spanish) to English...
Sys.setlocale('LC_TIME', 'en_US.UTF-8')

compressedFileName <- "repdata_data_StormData.csv.bz2"

URL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"


# If file doesn't exist in current working directory, I try to download from website
if(!file.exists(compressedFileName))
    download.file(URL, compressedFileName)

# If the file doesn't exist in current working directory after downloading, there are any
# problem downloading this file, or there are any other problem...aborting
if(!file.exists(compressedFileName))    
    stop("Required file doesn't exist in the current working directory! Aborting...")
    
# loading data
data <- read.csv(compressedFileName)

# Get the column names and check if they are properly formatted
colnames(data) <- make.names(colnames(data))

```

```{r echo=TRUE}

head(data)
```

### 2. Storm Data - Exploratory analysis 

There is some documentation of this dataset available [here](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf). 
More information can be obtained from the [NOAA Storm Events Database](http://www.ncdc.noaa.gov/stormevents/ftp.jsp). 

There are **`r nrow(data)` rows** or observations and **`r ncol(data)` columns** or variables in this dataset.

Reading these two websites above and, for the topic of this paper, we can subset the *data* CSV file extracting only the following columns or variables:   

* BGN_DATE (event start date)
* STATE (the USA state where event occurs)
* EVTYPE (climatic event)
* FATALITIES (numerical value for an occurrence of death)
* INJURIES (numerical value for non fatal damage, hurt, harm, ...)
* PROPDMG (numerical value for property damage)
* PROPDMGEXP (exponent -symbol/value- for adjusting PROPDMG value)
* CROPDMG (numerical value for crop damage)
* CROPDMGEXP (exponent -symbol/value- for adjusting CROPDMG value)

The others columns or variables aren't important for the analysis in this article.

Now, we are going to subset the dataset.

```{r echo=TRUE, cache=TRUE, message=FALSE, warning=FALSE}
suppressPackageStartupMessages(library(dplyr))

# select only the variables we are interested in
sub_data <- data %>% select(BGN_DATE, STATE, EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP,CROPDMG, CROPDMGEXP)

# For saving computer resources, we can remove the initial dataset
rm(data)

# now, we have a smaller dataset
head(sub_data)
```
At this moment, There are **`r nrow(sub_data)` rows** or observations and **`r ncol(sub_data)` columns** or variables in *sub_data* dataset.
For this analysis we want to study types of weather events that are most harmful with respect to population health and with respect to economic consequences, so we want dataset observations with values over "0" (zero) in the following variables: FATALITIES, INJURIES, PROPDMG and CROPDMG.

```{r echo=TRUE, cache=TRUE}

sub_data <- sub_data %>% filter(FATALITIES > 0 | INJURIES > 0 | PROPDMG > 0 | CROPDMG > 0)

```

Now, there are **`r nrow(sub_data)` rows** or observations and **`r ncol(sub_data)` columns** or variables in *sub_data* dataset. We have reduced the data size significantly.

```{r echo=TRUE, cache=TRUE}
#summary of our dataset
summary(sub_data)

```

### 3. Storm Data - Cleaning the Data

In the official documentation, [here](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf), we can read that there are 48 event types (chapter 7, points from 7.1 to 7.48, or the Storm Data Event Table in the page 6 of the above PDF document), and we have `r length(levels(sub_data$EVTYPE))` event types in our dataset. We'll need to transform and reduce them. 

In first place, we are going to know how many ocurrences of each event type there are in our *sub_data* dataset.

```{r echo=TRUE, cache=TRUE}
summary(sub_data$EVTYPE)
```

In this summary, we can see that THUNDERSTORM WIND and other variations of the same climatic event are the more common (TSTM WIND, THUNDERSTORM WINDS, THUNDERSTORM WINDSS, THUNDERSTORM, TSTM WIND (G45), THUNDERSTORM WINDS/HAIL, WINDS, GUSTY WINDS). The idea is to group these events in one big group with name **THUNDERSTORM WIND**, like this event is named in the official documentation. We'll do the same process for others climatic events with similar variable name in our dataset. This way we reduce the dataset dimensions.

Also, there are duplicate event names, and the same event name in lowercase and in uppercase, so we'll transform all event names to uppercase, for grouping purpouses.

```{r echo=TRUE, cache=TRUE}

# Transform all event names to UPPERCASE. This transform the column "as.character".
sub_data <- mutate(sub_data, EVTYPE = as.factor(toupper(EVTYPE)))

```


We are going to group similar climatic events that can be repetitive or similar events, like THUNDERSTORM WIND or TSTM WIND, for reduce this dataset. For this purpouse, the following "groupingFactorLevels" function replaces the terms passed in the "arrayNames" parameter with the term passed in the "keyName" parameter.

```{r echo=TRUE, cache=TRUE}

groupingFactorLevels <- function(mydata, keyName, arrayNames) 
{
# searching by terms inside arrayNAmes parameter
toMatch <- arrayNames

# get the indices where there are matches
indices <- grep(paste(toMatch, collapse="|"), levels(mydata$EVTYPE))


# replacement matches with keyName parameter
levels(mydata$EVTYPE)[c(indices)] <- keyName

return(mydata)

}

```


We are going to use the above function for grouping the most important climatic events, that is, the most frequent climatic events (as you can see in: summary(sub_data$EVTYPE))

```{r echo=TRUE, cache=TRUE, message=FALSE, warning=FALSE}

toMatch <-c("TORNADO", "TORNADOES")
sub_data <- groupingFactorLevels(sub_data, c("TORNADO"), toMatch)

toMatch <- c("^THUNDERSTORM", "^TSTM", "^SEVERE THUNDERSTORM", "^THUNDERTORM WINDS", "^TSTM WIND", "TSTM WIND")
sub_data <- groupingFactorLevels(sub_data, c("THUNDERSTORM WIND"), toMatch)

toMatch <-c("^MARINE TSTM WIND", "^MARINE THUNDERSTORM WIND" )
sub_data <- groupingFactorLevels(sub_data, c("MARINE THUNDERSTORM WIND"), toMatch)

toMatch <-c("HURRICANE", "TYPHOON", "HURRICANE/TYPHOON")
sub_data <- groupingFactorLevels(sub_data, c("HURRICANE/TYPHOON"), toMatch)

toMatch <-c("TROPICAL STORM")
sub_data <- groupingFactorLevels(sub_data, c("TROPICAL STORM"), toMatch)


toMatch <-c("RIVER FLOOD", "URBAN/SMALL STREAM FLOOD", "URBAN FLOODING", "MAJOR FLOOD", 
            "URBAN FLOOD", "URBAN/SML STREAM FLD", "FLOOD/RAIN/WINDS", "^FLOODS$", 
            "FLOOD & HEAVY RAIN", "RIVER AND STREAM FLOOD")
sub_data <- groupingFactorLevels(sub_data, c("FLOOD"), toMatch)

toMatch <-c("FLASH FLOOD", "FLASH FLOODING", "FLOODING", "FLASH FLOODS", "FLOOD FLASH",
            "FLOOD/FLASH")
sub_data <- groupingFactorLevels(sub_data, c("FLASH FLOOD"), toMatch)

toMatch <-c("COASTAL FLOOD", "COASTAL", "BEACH EROSION", "EROSION/CSTL FLOOD", "CSTL")
sub_data <- groupingFactorLevels(sub_data, c("COASTAL FLOOD"), toMatch)


toMatch <-c("LIGHT FREEZING RAIN", "FREEZE", "FREEZING RAIN", "FROST/FREEZE", 
           "AGRICULTURAL FREEZE", "FREEZING RAIN/SNOW", "DAMAGING FREEZE", "FREEZING DRIZZLE", 
           "FROST", "SNOW FREEZING RAIN", "ICY", "FREEZING SPRAY")
sub_data <- groupingFactorLevels(sub_data, c("FROST/FREEZE"), toMatch)

toMatch <-c("^COLD$", "^WIND CHILD$", "$COLD/WIND CHILL$")
sub_data <- groupingFactorLevels(sub_data, c("COLD/WIND CHILL"), toMatch)

toMatch <-c("^EXTREME WIND CHILL", "RECORD COLD")
sub_data <- groupingFactorLevels(sub_data, c("EXTREME COLD/WIND CHILL"), toMatch)

toMatch <-c("SNOW") 
sub_data <- groupingFactorLevels(sub_data, c("HEAVY SNOW"), toMatch)

toMatch <-c("^HAIL") 
sub_data <- groupingFactorLevels(sub_data, c("HAIL"), toMatch)

toMatch <-c("MARINE HAIL") 
sub_data <- groupingFactorLevels(sub_data, c("MARINE HAIL"), toMatch)

# This new data type is to summarize... 
toMatch <-c("FOG") 
sub_data <- groupingFactorLevels(sub_data, c("DENSE/FREEZING FOG"), toMatch)

toMatch <-c("^HEAT") 
sub_data <- groupingFactorLevels(sub_data, c("HEAT"), toMatch)

toMatch <-c("^EXCESSIVE HEAT", "DROUGHT/EXCESSIVE HEAT", "RECORD HEAT") 
sub_data <- groupingFactorLevels(sub_data, c("EXCESSIVE HEAT"), toMatch)

#toMatch <-c("^DROUGHT") 
#sub_data <- groupingFactorLevels(sub_data, c("DROUGHT"), toMatch)

toMatch <-c("BLIZZARD") 
sub_data <- groupingFactorLevels(sub_data, c("BLIZZARD"), toMatch)

toMatch <-c("^EXTREME COLD", "EXTENDED COLD") 
sub_data <- groupingFactorLevels(sub_data, c("EXTREME COLD/WIND CHILL"), toMatch)

toMatch <-c("^COLD") 
sub_data <- groupingFactorLevels(sub_data, c("COLD/WIND CHILL"), toMatch)

toMatch <-c("^ICE STORM", "BLACK ICE", "GLAZE ICE", "ICE FLOES", "GLAZE/ICE STORM", "GLAZE") 
sub_data <- groupingFactorLevels(sub_data, c("ICE STORM"), toMatch)

toMatch <-c("^HIGH WIND", "^WIND", "GUSTY WIND", "GRADIENT WIND") 
sub_data <- groupingFactorLevels(sub_data, c("HIGH WIND"), toMatch)

toMatch <-c("^STRONG WIND") 
sub_data <- groupingFactorLevels(sub_data, c("STRONG WIND"), toMatch)

toMatch <-c("DUST DEVIL") 
sub_data <- groupingFactorLevels(sub_data, c("DUST DEVIL"), toMatch)

toMatch <-c("^DUST$", "BLOWING DUST") 
sub_data <- groupingFactorLevels(sub_data, c("DUST"), toMatch)

toMatch <-c("^RAIN", "HEAVY PRECIPITATION", "^HEAVY RAIN", "RAINFALL", "UNSEASONAL RAIN", "RAIN", "PRECIP") 
sub_data <- groupingFactorLevels(sub_data, c("HEAVY RAIN"), toMatch)

toMatch <-c("SMOKE") 
sub_data <- groupingFactorLevels(sub_data, c("DENSE SMOKE"), toMatch)

toMatch <-c("AVALACE") 
sub_data <- groupingFactorLevels(sub_data, c("AVALANCHE"), toMatch)



summary(sub_data$EVTYPE)
```

The most important climatic events are summarized in the above list, and for the purpouse of this article, it is enough.

### 4. Storm Data - Summarizing the Data

Now, we have two different questions to resolve:

> 1. Weather events that are most harmful with respect to population health
> 2. Weather events that are most harmful with respect to economic consequences

The climatic events that affect the first question may not be the same that affect the second question and vice versa.

For this purpouse, we need to extract these differents observations from the *sub_data* dataset. One dataset for population health, and another dataset for economic consequences.

```{r echo=TRUE, cache=TRUE}
sub_data_health <- sub_data %>% filter(FATALITIES > 0 | INJURIES > 0)
sub_data_economic <- sub_data %>% filter(PROPDMG > 0 | CROPDMG > 0)

# For saving computer resources, we can remove the initial dataset
rm(sub_data)
```

Now, there are **`r nrow(sub_data_health)` rows** or observations and **`r ncol(sub_data_health)` columns** or variables in *sub_data_health* dataset, and  **`r nrow(sub_data_economic)` rows** or observations and **`r ncol(sub_data_economic)` columns** or variables in *sub_data_economic* dataset.

At this point, we are going to check the more harmful event types in our datasets.

```{r echo=TRUE, cache=TRUE}

# Group by event type
my_health_group <- group_by(sub_data_health, EVTYPE)

# Get the sum of injuries and fatalities by event type

my_health_summary_fatalities <- summarise(my_health_group, sum(FATALITIES))
my_health_summary_injuries <- summarise(my_health_group, sum(INJURIES))
my_health_total_summary <- summarise(my_health_group, sum(FATALITIES),sum(INJURIES), 
                                     TOTAL = sum(FATALITIES) + sum(INJURIES))

# change tne column names
colnames(my_health_summary_fatalities) <- c("EVTYPE", "SUM_FATALITIES")
colnames(my_health_summary_injuries) <- c("EVTYPE", "SUM_INJURIES")
colnames(my_health_total_summary) <- c("EVTYPE", "SUM_FATALITIES", "SUM_INJURIES", "TOTAL")

# Order the dataset by number of FATALITIES & INJURIES
my_health_summary_fatalities <- arrange(my_health_summary_fatalities, desc(SUM_FATALITIES))
my_health_summary_injuries <- arrange(my_health_summary_injuries, desc(SUM_INJURIES))
my_health_total_summary <- arrange(my_health_total_summary, desc(TOTAL))



```


And now, we are going to check the more harmful event types with respect to economic consequences. Our dataset has four fields:

* PROPDMG (numerical value for property damage)
* PROPDMGEXP (exponent -symbol/value- for adjusting PROPDMG value)
* CROPDMG (numerical value for crop damage)
* CROPDMGEXP (exponent -symbol/value- for adjusting CROPDMG value)

PROPDMGEXP and CROPDMGEXP fields are factor variables with the following values:

```{r echo=TRUE, cache=TRUE}

levels(sub_data_economic$PROPDMGEXP)
levels(sub_data_economic$CROPDMGEXP)

```

We need to tranform the PROPDMGEXP and CROPDMGEXP to a numerical value and to multiply by PROPDMG and CROPDMG values, respectively. In this way, we will have the damage value. The next table shows the levels of the factor variable and their numerical values associated:

                     * ""  = 1 
                     * "-" = 1
                     * "?" = 1
                     * "+" = 1
                     * "0" = 1
                     * "1" = 10
                     * "2" = 100
                     * "3" = 1000
                     * "4" = 10000
                     * "5" = 100000
                     * "6" = 1000000
                     * "7" = 10000000
                     * "8" = 100000000
                     * "9" = 1000000000
                     * "h" = 100
                     * "H" = 100
                     * "k" = 1000
                     * "K" = 1000
                     * "m" = 1000000
                     * "M" = 1000000
                     * "b" = 1000000000
                     * "B" = 1000000000
                     

```{r echo=TRUE, cache=TRUE}

# > levels(sub_data_economic$PROPDMGEXP)
# [1] ""  "-" "?" "+" "0" "1" "2" "3" "4" "5" "6" "7" "8" "B" "h" "H" "K" "m" "M"

levels(sub_data_economic$PROPDMGEXP) <- c(1,1,1,1,1,
                                          10,100,1000,10000,100000,1000000,
                                          10000000,100000000,1000000000,
                                          100, 100, 1000, 1000000, 1000000)


# > levels(sub_data_economic$CROPDMGEXP)
# [1] ""  "?" "0" "2" "B" "k" "K" "m" "M"

levels(sub_data_economic$CROPDMGEXP) <- c(1,1,1,100,1000000000,1000,1000,1000000,1000000)

# Show the new levels
levels(sub_data_economic$PROPDMGEXP)
levels(sub_data_economic$CROPDMGEXP)

```

Now, we can calculate the numerical crop and properties damage in dollars.

```{r echo=TRUE, cache=TRUE}

# Add two new variables with the total damage, PROPDMG_TOTAL and CROPDMG_TOTAL
sub_data_economic <- sub_data_economic %>% 
                     mutate(PROPDMG_TOTAL = PROPDMG * as.numeric(as.character(PROPDMGEXP)), 
                            CROPDMG_TOTAL = CROPDMG * as.numeric(as.character(CROPDMGEXP)))


# Group by event type
my_economic_group <- group_by(sub_data_economic, EVTYPE)

# Get the sum of injuries and fatalities by event type
my_economic_summary_propdmg <- summarise(my_economic_group, sum(PROPDMG_TOTAL))
my_economic_summary_cropdmg <- summarise(my_economic_group, sum(CROPDMG_TOTAL))
my_economic_total_summary <- summarise(my_economic_group, sum(PROPDMG_TOTAL),sum(CROPDMG_TOTAL),
                                     TOTAL = sum(PROPDMG_TOTAL) + sum(CROPDMG_TOTAL))

# change tne column names
colnames(my_economic_summary_propdmg) <- c("EVTYPE", "SUM_PROPDMG_TOTAL")
colnames(my_economic_summary_cropdmg) <- c("EVTYPE", "SUM_CROPDMG_TOTAL")
colnames(my_economic_total_summary) <- c("EVTYPE", "SUM_PROPDMG_TOTAL", 
                                          "SUM_CROPDMG_TOTAL", "TOTAL_DAMAGE")

# Order the dataset by number of total PROP. DAMAGE & total CROP DAMAGE
my_economic_summary_propdmg <- arrange(my_economic_summary_propdmg, desc(SUM_PROPDMG_TOTAL))
my_economic_summary_cropdmg <- arrange(my_economic_summary_cropdmg, desc(SUM_CROPDMG_TOTAL))
my_economic_total_summary <- arrange(my_economic_total_summary, desc(TOTAL_DAMAGE))



```

# **Results**

### 1. Show the final data and plotting Health results and conclusions

```{r echo=TRUE }
# Preview
head(my_health_summary_fatalities)
head(my_health_summary_injuries)
head(my_health_total_summary)
```



```{r echo=TRUE, cache=TRUE}
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(grid))
suppressPackageStartupMessages(library(gridExtra))

#Subsetting the six most relevant events
x <- my_health_summary_fatalities[1:6, ]
y <- my_health_summary_injuries[1:6, ]
z <- my_health_total_summary[1:6, ]

x$EVTYPE <- reorder(x$EVTYPE, x$SUM_FATALITIES)

# Create a first plot for the fatalities dataset 
p <- ggplot(x, aes(x=EVTYPE, y=SUM_FATALITIES, fill=SUM_FATALITIES))
p <- p + geom_histogram(stat='identity') + coord_flip()
p <- p + scale_fill_gradient("Nº Fatalities", 
                             low = "green", 
                             high = "red")
 # p <- p + theme(text = element_text(size=10),axis.text.x = element_text(angle=90, vjust=1))
 p <- p + theme(text = element_text(size=8)) 
 p <- p + xlab("Event Type")
 p <- p + ylab("Total Fatalities")
 p <- p + ggtitle("Total Fatalities By Event Type")
 
 
 y$EVTYPE <- reorder(y$EVTYPE, y$SUM_INJURIES)

# Create a second plot for the injuries dataset 
p2 <- ggplot(y, aes(x=EVTYPE, y=SUM_INJURIES, fill=SUM_INJURIES))
p2 <- p2 + geom_histogram(stat='identity') + coord_flip()
p2 <- p2 + scale_fill_gradient("Nº Injuries", 
                             low = "green", 
                             high = "red")
 # p <- p + theme(text = element_text(size=10),axis.text.x = element_text(angle=90, vjust=1))
 p2 <- p2 + theme(text = element_text(size=8)) 
 p2 <- p2 + xlab("Event Type")
 p2 <- p2 + ylab("Total Injuries")
 p2 <- p2 + ggtitle("Total Injuries By Event Type")
 
 
 z$EVTYPE <- reorder(z$EVTYPE, z$TOTAL)

# Create a third plot for the TOTAL health dataset 
p3 <- ggplot(z, aes(x=EVTYPE, y=TOTAL, fill=TOTAL))
p3 <- p3 + geom_histogram(stat='identity') + coord_flip()
p3 <- p3 + scale_fill_gradient("Nº Total", 
                             low = "green", 
                             high = "red")
 # p <- p + theme(text = element_text(size=10),axis.text.x = element_text(angle=90, vjust=1))
 p3 <- p3 + theme(text = element_text(size=8)) 
 p3 <- p3 + xlab("Event Type")
 p3 <- p3 + ylab("Total Fatalities + Injuries")
 p3 <- p3 + ggtitle("Total Fatalities + Injuries By Event Type")
 
 
 ## Call to multipot function for plotting the 2 plots in one column
 # multiplot(p, p3, p2, cols = 2)
 # multiplot(p, p2, p3, matrix(c(1,2,3,3), nrow=1, byrow=TRUE))

 # Get the gtables
gA <- ggplotGrob(p)
gB <- ggplotGrob(p2)
gC <- ggplotGrob(p3)

grid.arrange(gA, gB, gC, nrow=3)

```

As a conclusion we can see clearly the TORNADO as the event type that is most harmful with respect to population health, followed by HEAT/EXCESSIVE HEAT and THUNDERSTORM WIND. 

### 2. Show the final data and plotting Economic results and conclusions


```{r echo=TRUE}
# Preview
head(my_economic_summary_propdmg)
head(my_economic_summary_cropdmg)
head(my_economic_total_summary)


```



```{r echo=TRUE, cache=TRUE}
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(grid))
suppressPackageStartupMessages(library(gridExtra))

#Subsetting the six most relevant events
x <- my_economic_summary_propdmg[1:6, ]
y <- my_economic_summary_cropdmg[1:6, ]
z <- my_economic_total_summary[1:6, ]

x$EVTYPE <- reorder(x$EVTYPE, x$SUM_PROPDMG_TOTAL)

# Create a first plot for the properties damage dataset 
p <- ggplot(x, aes(x=EVTYPE, y=SUM_PROPDMG_TOTAL, fill=SUM_PROPDMG_TOTAL))
p <- p + geom_histogram(stat='identity') + coord_flip()
p <- p + scale_fill_gradient("USD", 
                             low = "green", 
                             high = "red")
 # p <- p + theme(text = element_text(size=10),axis.text.x = element_text(angle=90, vjust=1))
 p <- p + theme(text = element_text(size=8)) 
 p <- p + xlab("Event Type")
 p <- p + ylab("Total Property Damage (USD)")
 p <- p + ggtitle("Total Property Damage (USD) By Event Type")
 
 
 y$EVTYPE <- reorder(y$EVTYPE, y$SUM_CROPDMG_TOTAL)

# Create a second plot for the crop damage dataset 
p2 <- ggplot(y, aes(x=EVTYPE, y=SUM_CROPDMG_TOTAL, fill=SUM_CROPDMG_TOTAL))
p2 <- p2 + geom_histogram(stat='identity') + coord_flip()
p2 <- p2 + scale_fill_gradient("USD", 
                             low = "green", 
                             high = "red")
 # p <- p + theme(text = element_text(size=10),axis.text.x = element_text(angle=90, vjust=1))
 p2 <- p2 + theme(text = element_text(size=8)) 
 p2 <- p2 + xlab("Event Type")
 p2 <- p2 + ylab("Total Crop Damage (USD)")
 p2 <- p2 + ggtitle("Total Crop Damage By Event Type")
 
 
 z$EVTYPE <- reorder(z$EVTYPE, z$TOTAL_DAMAGE)

# Create a third plot for the TOTAL economic dataset 
p3 <- ggplot(z, aes(x=EVTYPE, y=TOTAL_DAMAGE, fill=TOTAL_DAMAGE))
p3 <- p3 + geom_histogram(stat='identity') + coord_flip()
p3 <- p3 + scale_fill_gradient("USD", 
                             low = "green", 
                             high = "red")
 # p <- p + theme(text = element_text(size=10),axis.text.x = element_text(angle=90, vjust=1))
 p3 <- p3 + theme(text = element_text(size=8)) 
 p3 <- p3 + xlab("Event Type")
 p3 <- p3 + ylab("Total Properties + Crop Damage")
 p3 <- p3 + ggtitle("Total Properties + Crop Damage By Event Type")
 
 
 ## Call to multipot function for plotting the 2 plots in one column
 # multiplot(p, p3, p2, cols = 2)
 # multiplot(p, p2, p3, matrix(c(1,2,3,3), nrow=1, byrow=TRUE))

 # Get the gtables
gA <- ggplotGrob(p)
gB <- ggplotGrob(p2)
gC <- ggplotGrob(p3)

grid.arrange(gA, gB, gC, nrow=3)

```

As a conclusion we can see clearly the FLOOD as the event type that is most harmful with respect to economic consequences, followed by HURRICANE/TYPHOON and TORNADO. Although the DROUGHT event is the most harmful with respect to CROP damage.
